Appendix A

### 2023-11-14 manuscript/manuscript.Rmd
Work on manuscript revision.

For both invariant transformations, $\lambda$, $x_0$ and $s$ are obtained by maximising the log-likelihood function. A full derivation is shown in Appendix A. The invariant Box-Cox log-likelihood function is:

```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu$ and $\sigma^2$ are the mean and variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively. Similarly, for the invariant Yeo-Johnson log-likelihood function is:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-shift-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu$ and $\sigma^2$ are the mean and variance of the Box-Cox transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

@@

Real-world data may contain outliers, to which maximum likelihood estimation can be sensitive. Their presence may lead to poor transformations to normality, as shown in Figure \ref{fig:sensitivity-to-outliers}. As indicated by @Raymaekers2021-kq, the general aim of power transformations should be to transform non-outlier data to normality, i.e. achieve *central normality*. To achieve this, they devised an iterative procedure to find a robust estimate of the transformation parameter $\lambda$. Because $\lambda$ is updated during their procedure, simultaneous estimation of $\lambda$ and $x_0$ for shift-sensitive power transformations is not possible.

In essence, obtaining robust transformation parameters that provide central normality requires identifying outliers in the data and weighting such instances during the optimisation process. @Raymaekers2021-kq do this (during the final steps of their iterative procedure) through a weighted maximum likelihood estimation. This weighted maximum log-likelihood estimator is based on equations \ref{eqn:box-cox-shift-log-likelihood} and \ref{eqn:yeo-johnson-shift-log-likelihood}. Here, the estimator is further normalised by the total weight and extended to include the shift parameter. The shift-sensitive Box-Cox power transformation is:

```{=latex}
\begin{equation}
\left\{ \hat{\lambda}^{\text{rBC}}, \hat{x}_0^{\text{rBC}} \right\} = \argmax_{\lambda, x_0} \frac{1}{\sum_{i=1}^n w_i} \sum_{i=1}^n w_i \left[ -\frac{1}{2}\log(\sigma^2_w) + (\lambda - 1) \log(x_i - x_0) \right]
\end{equation}
```
where $\sigma^2_w$ is the weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n \phi_{\text{BC}}^{\lambda, x_0} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```
Analogously, the weighted maximum log-likelihood estimator for the shift-sensitive Yeo-Johnson power transformation is:

```{=latex}
\begin{equation}
\left\{ \hat{\lambda}^{\text{rYJ}}, \hat{x}_0^{\text{rYJ}} \right\} = \argmax_{\lambda, x_0} \frac{1}{\sum_{i=1}^n w_i} \sum_{i=1}^n w_i \left[ -\frac{1}{2}\log(\sigma^2_w) + (\lambda - 1) \sgn(x_i - x_0) \log(1 + |x_i - x_0|) \right]
\end{equation}
```
where $\sigma^2_w$ is the weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n \phi_{\text{YJ}}^{\lambda, x_0} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```

Other criteria aside from MLE can be used for finding robust transformation parameters, see the Appendix. For the remainder, we will use MLE as the optimisation criterion.

@@

- Set weights based on the z-score of the transformed feature $\phi^{\lambda, x_0} (\mathbf{X})$. After @Raymaekers2021-kq, $z_i = \frac{\phi^{\lambda, x_0}(x_i) - \mu_M}{\sigma_M}$. Here, $\mu_M$ and $\sigma_M$ are robust Huber M-estimates of location and scale of the transformed feature $\phi^{\lambda, x_0} (\mathbf{X})$ [@Huber1981-su].

- Set weights based on the residual error between the z-score of the transformed feature $\phi^{\lambda, x_0} (\mathbf{X})$ and the theoretical z-score from a standard normal distribution: $r_i =\left| \left( \phi^{\lambda, x_0}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|$, with $\mu_M$, $\sigma_M$ and $p_i$ as defined above.

>>>>

For both invariant transformations, $\lambda$, $x_0$ and $s$ parameters are obtained by maximising the log-likelihood function. A full derivation is shown in Appendix A. The invariant Box-Cox log-likelihood function is:

```{=latex}
\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu$ and $\sigma^2$ are the mean and variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively. Similarly, for the invariant Yeo-Johnson log-likelihood function is:

```{=latex}
\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu$ and $\sigma^2$ are the mean and variance of the Box-Cox transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

@@

Real-world data may contain outliers, to which maximum likelihood estimation can be sensitive. Their presence may lead to poor transformations to normality, as shown in Figure \ref{fig:sensitivity-to-outliers}. As indicated by @Raymaekers2021-kq, the general aim of power transformations should be to transform non-outlier data to normality, i.e. achieve *central normality*. To achieve this, they devised an iterative procedure to find a robust estimate of the transformation parameter $\lambda$. Because $\lambda$ is updated during their procedure, it can not be used simultaneously estimate $\lambda$, $x_0$ and $s$ for invariant power transformations.

In essence, obtaining robust transformation parameters that provide central normality requires identifying outliers in the data and weighting such instances during the optimisation process. @Raymaekers2021-kq do this (during the final steps of their iterative procedure) through a weighted maximum likelihood estimation. Here, weighted maximum likelihood estimation is based on equations \ref{eqn:box-cox-invariant-log-likelihood} and \ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to @Raymaekers2021-kq, these log-likelihood functions includes additional terms to accommodate estimation of $x_0$ and $s$. The weighted invariant Box-Cox log-likelihood function is:

```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n \phi_{\text{BC}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```

Analogously, the weighted invariant Yeo-Johnson log-likelihood function is:

```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$:

```{=latex}
\begin{equation}
\sigma_w^2 = \frac{\sum_{i=1}^n w_i \left(\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) - \mu_w \right)^2}{\sum_{i=1}^n w_i} \quad \text{with } \mu_w = \frac{\sum_{i=1}^n \phi_{\text{YJ}}^{\lambda, x_0, s} (x_i)} {\sum_{i=1}^n w_i}
\end{equation}
```

Other criteria aside from MLE can be used for finding robust transformation parameters, see Appendix B. For the remainder, we will use MLE as the optimisation criterion.

@@

- Set weights based on the z-score of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$. After @Raymaekers2021-kq, $z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}$. Here, $\mu_M$ and $\sigma_M$ are robust Huber M-estimates of location and scale of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ [@Huber1981-su].

- Set weights based on the residual error between the z-score of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ and the theoretical z-score from a standard normal distribution: $r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|$, with $\mu_M$, $\sigma_M$ and $p_i$ as defined above.

### 2023-11-14 manuscript/manuscript_appendix_a.Rmd
Work on manuscript revision.

We introduced per-sample weights $w_i$ to facilitate robust power transformations. The log likelihood of the robust Box-Cox transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\sigma^2_w$ is the weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$:

@@

The log likelihood of the robust Yeo-Johnson transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\sigma^2_w$ is the weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{BC}}^{\lambda, x_0} (\mathbf{X})$:

>>>>

We introduced per-sample weights $w_i$ to facilitate robust power transformations. The log likelihood of the robust Box-Cox transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively:

@@

The log likelihood of the robust Yeo-Johnson transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$:
